---
title: "Analysis"
author: "Lily Zhu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(nnet)
```

```{r}
us_19 <- read_csv('~/df_data/US/us_19.csv')
```

## Data Transformation

**Factor variables**

``` {r}
col_names <- colnames(us_19)
us_19 <- lapply(us_19, as.factor)
us_19 <- data.frame(matrix(unlist(us_19), nrow=length(us_19), byrow=TRUE))
us_19 <- data.frame(t(us_19))
colnames(us_19) <- col_names
```

**Distribution of response variable**

```{r dist of response, fig.height = 3, fig.width = 5, fig.align='center'}
ggplot(data = us_19, aes(x = DAST_CAT))+
  geom_bar(fill = "#00539B") +
  labs(title = "Distribution of DAST-10 Score Category",
       x = "DAST-10 Score Category", y = "Count") +
  theme_bw()
```

```{r response stats}
us_19 %>% 
  summarise(mean = mean(as.numeric(DAST_CAT)),
          median = median(DAST_CAT),
          sd = sd(DAST_CAT),
          iqr = IQR(DAST_CAT),
          min = min(DAST_CAT),
          max = max(DAST_CAT))
```

**Distribution of predictor variables**

```{r dist of predictors, fig.height = 3, fig.width = 6, fig.align='center'}
painrel <- ggplot(data = us_19, aes(x = PAINREL_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription pain relievers") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

stim <- ggplot(data = us_19, aes(x = STIM_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription stimulants") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

sed <- ggplot(data = us_19, aes(x = SED_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription sedatives") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

thc <- ggplot(data = us_19, aes(x = THC_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription cannabinoids") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

op <- ggplot(data = us_19, aes(x = OP_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription opioids") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

gaba <- ggplot(data = us_19, aes(x = GABA_NMUYR))+
  geom_bar(fill = "#00539B") +
   xlab("NMU in last 12 months") + ylab("Count") +
   ggtitle("Prescription GABA-analogues") +
   scale_x_discrete(labels=c("0" = "No", "1" = "Yes")) +
   theme_bw(base_size = 9) +
   theme(plot.title = element_text(size = 9, hjust = 0.5))

(painrel | stim | sed) /
  (thc | op | gaba)
```

**Relationship between predictor variables and response variable**

```{r fig.height = 3, fig.width = 6, fig.align='center'}
painrel_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = PAINREL_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription pain reliever \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

stim_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = STIM_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription stimulant \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

sed_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = SED_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription sedative \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

thc_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = THC_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription cannabinoid \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

op_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = OP_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription opioid \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

gaba_dast_10 <- ggplot(data = us_19, mapping = aes(x = DAST_CAT, fill = GABA_NMUYR)) +
  geom_bar(position = "fill") + 
  labs(title = "Prescription GABA-analogue \nlifetime use", 
       x = "DAST-10 category", y = "") +
  theme_bw(base_size = 9) +
  theme(plot.title = element_text(size = 9, hjust = 0.5)) +
  scale_fill_manual(values=c("#00539B", "lightblue"), name = "", labels = c("No", "Yes"), guide = guide_legend(reverse=TRUE))

(painrel_dast_10 + stim_dast_10) /
(sed_dast_10 + thc_dast_10)
```

```{r fig.height = 1.5, fig.width = 6, fig.align='center'}
op_dast_10 + gaba_dast_10
```

## Model Selection

**Multinomial Model**

```{r}
set.seed(2021)
us_19_mod <- us_19 %>%
  sample_n(1000) %>% 
  mutate(DAST_binary = if_else(DAST_SUM < 3, 0, 1))
```

```{r}
# us_19_mod %>% 
#      select(starts_with("DEM"))

us_19_mod <- us_19_mod %>% 
  select(ends_with("NMUYR"), starts_with("DEM"), TOB_LIFE, DAST_CAT) %>% 
  subset(select = -c(DEM_PREG, DEM_PREG_WK, DEM_STDNT_PROGRAM, DEM_STDNT_UNDER, DEM_VET_SERV, DEM_HEALTH_PRIV, DEM_HEALTH_PRE, DEM_HEALTH_HOS, DEM_HEALTH_OUT, DEM_HEALTH_HOME, DEM_HEALTH_OTH, DEM_HEALTH_ROLE))
```

```{r full model, results = "hide"}
full_model <- multinom(DAST_CAT ~ PAINREL_NMUYR + STIM_NMUYR + SED_NMUYR + THC_NMUYR + OP_NMUYR + GABA_NMUYR + DEM_GENDER + DEM_AGE10 + DEM_REGION + DEM_INCOME + TOB_LIFE, 
                       data = us_19_mod)
```

```{r tidy full model}
# tidy(full_model) %>%
#   kable(digits = 3)
```

```{r backward model selection, results = "hide"}
int_only_model <- multinom(DAST_CAT ~ 1, data = us_19_mod)

final_model <- step(full_model, scope = formula(int_only_model), direction = "backward")
```

```{r tidy final model}
tidy(final_model, conf.int = TRUE) %>% 
  kable(digits = 3)
```

**Binary Model**

```{r binary model}
binary_model <- glm(DAST_binary ~ PAINREL_NMUYR + STIM_NMUYR + SED_NMUYR + THC_NMUYR + OP_NMUYR + GABA_NMUYR + DEM_GENDER + DEM_AGE10 + DEM_REGION + DEM_INCOME + TOB_LIFE, data = us_19_mod, family = binomial)
```

```{r tidy binary model}
tidy(binary_model) %>% 
  kable(digits = 3)
```

```{r backward binary model selection, results = "hide"}
int_only_binary_model <- glm(DAST_binary ~ 1, data = us_19_mod, family = binomial)

final_binary_model <- step(binary_model, scope = formula(int_only_binary_model), direction = "backward")
```

```{r tidy final binary model}
tidy(final_binary_model, conf.int = TRUE) %>% 
  kable(digits = 3)
```

# Results

## Model Assumptions

```{r}
# us_19_mod %>%
#   count(PAINREL_NMUYR, DAST_binary) %>%
#   group_by(PAINREL_NMUYR) %>%
#   mutate(prop = n/sum(n)) %>%
#   filter(DAST_binary == "1") %>%
#   mutate(emp_logit = log(prop/(1-prop)))
```

```{r}
# library(Stat2Data)
# emplogitplot2(DAST_binary ~ PAINREL_NMUYR + STIM_NMUYR + THC_NMUYR + DEM_GENDER + DEM_AGE + DEM_INCOME + TOB_LIFE, data = us_19_mod)
```

The randomness condition is satisfied. We do not have reason to believe that the participants in this study differ systematically from adults in the U.S.

The independence condition is satisfied. It is reasonable to conclude that the participants' health and drug use behavior characteristics are independent of one another.

## Model Diagnostics

**Leverage**

```{r leverage threshold}
#calculate threshold
leverage_threshold <- 2*(16+1)/1000
```

```{r leverage v obs num, fig.height = 3, fig.width = 5, fig.align='center'}
##add row number to help with graphing 
final_model_augmented <- augment(final_binary_model) %>%
   mutate(obs_num = row_number())

ggplot(data = final_model_augmented, aes(x = obs_num, y = .hat)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = leverage_threshold, color = "red") +
  labs(x = "Observation Number", y = "Leverage") +
  geom_text(aes(label=ifelse(.hat > leverage_threshold,
                             as.character(obs_num), "")), nudge_x = 4) +
  theme_bw()
```

**Standardized Residuals**

```{r std resid, fig.height = 3, fig.width = 5, fig.align='center'}
#scatterplot of std resid vs predicted
ggplot(data = final_model_augmented, aes(x = obs_num, y = .std.resid)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 0, color = "red") +
  geom_hline(linetype = "dotted", yintercept = c(-2,2)) +
  geom_hline(linetype = "dotted", yintercept = c(-3,3), color = "red") +
  labs(x = "Observation Number", y = "Standardized Residuals") +
  geom_text(aes(label=ifelse(.std.resid > 3 | .std.resid < -3,
                             as.character(obs_num), "")), nudge_x = 4) +
  theme_bw()
```

```{r potential outlier}
final_model_augmented %>% 
  filter(abs(.std.resid) > 3) %>%
  kable(digit = 3)
```

**Cook's Distance**

```{r cooks d, fig.height = 3, fig.width = 5, fig.align='center'}
#scatterplot of cook's d vs obs num
ggplot(data = final_model_augmented, aes(x = obs_num, y = .cooksd)) +
  geom_point(alpha = 0.7) +
  geom_hline(yintercept = 1, color = "red") +
  geom_hline(linetype = "dotted", yintercept = 0.5, color = "red") +
  labs(x = "Observation Number", y = "Cook's D") +
  geom_text(aes(label=ifelse(.cooksd > 0.5,
                             as.character(obs_num), "")), nudge_x = 4) +
  theme_bw()
```

**Multicollinearity**

```{r check multicollinearity}
vif(final_binary_model) %>%
  tidy() %>%
  kable(digits = 3)
```

## Interpretation of Final Model, Discussion & Conclusion

```{r}
glance(final_binary_model) %>% 
  select(AIC) %>%
  kable(digit = 3, caption = "Final Binary Model")
```

```{r}
glance(binary_model) %>% 
  select(AIC) %>%
  kable(digit = 3, caption = "Binary Model (before backward selection)")
```
